\documentclass[oneside,a4paper,11pt]{article} 
\usepackage{fontspec}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xltxtra} 
\usepackage{polyglossia} 
\usepackage[table]{xcolor}
\usepackage{gb4e} 
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{float}
\usepackage{lineno}
\usepackage{textcomp}
\usepackage{hyperref} 
\hypersetup{bookmarks=false,bookmarksnumbered,bookmarksopenlevel=5,bookmarksdepth=5,xetex,colorlinks=true,linkcolor=blue,citecolor=blue}
\usepackage[all]{hypcap}
\usepackage{memhfixc}
\usepackage{lscape}
 \usepackage{setspace}

%\setmainfont[Mapping=tex-text,Numbers=OldStyle,Ligatures=Common]{Charis SIL} 
\newfontfamily\phon[Mapping=tex-text,Ligatures=Common,Scale=MatchLowercase,FakeSlant=0.3]{Charis SIL} 
\newcommand{\ipa}[1]{{\phon#1}} %API tjs en italique
 
\newcommand{\grise}[1]{\cellcolor{lightgray}\textbf{#1}}
\newfontfamily\cn[Mapping=tex-text,Ligatures=Common,Scale=MatchUppercase]{MingLiU}%pour le chinois
\newcommand{\zh}[1]{{\cn #1}}
\newcommand{\topic}{\textsc{dem}}
\newcommand{\tete}{\textsuperscript{\textsc{head}}}
\newcommand{\rc}{\textsubscript{\textsc{rc}}}
\XeTeXlinebreaklocale 'zh' %使用中文换行
\XeTeXlinebreakskip = 0pt plus 1pt %
 %CIRCG
\newcommand{\ro}{$\Sigma$}
\newcommand{\siga}{$\Sigma_1$} 
\newcommand{\sigc}{$\Sigma_3$}   
\newfontfamily\mleccha[Mapping=tex-text,Ligatures=Common,Scale=MatchLowercase]{Galatia SIL}%pour le grec
\newcommand{\grec}[1]{{\mleccha #1}}
\linenumbers

\begin{document} 

\title{A counterexample to Watkins' law in Siouan\footnote{Glosses follow the Leipzig Glossing rules, except for \textsc{prox} proximate (on which see \citealt{eschenberg05omaha}). The underscore \_ in a verb form indicates where the personal prefixes are inserted, in the case of verbs with discontinuous stems. In Lakota, capital \ipa{A} indicates an ablauting final vowel, which is realized as \ipa{a}, \ipa{e} or \ipa{iŋ} depending on the present of particular enclitics. Acknowledgements will be added after editorial decision.}}
 %Acknowledgements: Anton Antonov
%\author{Guillaume Jacques}
\maketitle
 \onehalfspacing
\textbf{Abstract}: This paper documents a case of analogy from the first person to the third in the Dhegiha languages. It discusses the significance of this example for historical linguistics in general, and proposes that higher frequency in discourse of the first person form in the case of cognition verbs explains why analogy took place in the opposite direction in the case of this verb.

\textbf{Keywords}: Siouan, Lakota, Dhegiha, Omaha, Osage, defective verb, analogy, verb of cognition, sound laws
 
\section{Introduction}
% \raggedright

 
In the Neogrammarian model, while exceptionless sound laws are fully deterministic, the same cannot be said of analogy, by definition irregular and unpredictable. Yet, it is possible to some extent to constrain  the space of hypotheses involving analogy, and research on the general principles of analogy is of utmost importance for historical linguistics.

The most successful observation involving analogical change is without doubt \citet{watkins62celtic}'s law, according to which the third person (especially singular) form is always the basis for analogical levelling in personal paradigms; the third person can be reanalyzed as being zero marked, and the resulting stem is generalized to the whole paradigm. The validity of this rule has been tested in various language families, and is apparent even in direct / inverse systems, where analogy operates from third person to SAP forms and first affects plural forms (see \citealt{jacques15directionality}, with examples from Algonquian).
    

Counterexamples to the directionality from third to first and second person in paradigms is not attested in Indo-European, Semitic or Algonquian languages. The only known cases where a first or second person form resists analogy better than the third person singular form occur when a phonological alternation is levelled across the board. For instance in Attic Greek,  where  the labiovelars become dentals before front vowels and labials in most other contexts (\citealt[253]{fortson10intro}), most verbal and nominal paradigms should present alternations between dental and labial stops in words whose last stem consonant was a labiovelar. A labial consonant would have been expected before endings in non-front vowels (for instance in the  first person singular \ipa{-ō} of the present thematic paradigm), and  dentals before front vowels (for instance the third person singular \ipa{-ei}). Yet, no  verb shows such alternation, as the labiovelars have been generalized everywhere:  both the first person singular \grec{νήφω} \ipa{nḗpʰō}  $\leftarrow $ *\ipa{ne-egʰʷ-ō} `I am sober' and the third person singular \grec{νήφει} \ipa{nḗpʰei}  $\leftarrow $ *\ipa{ne-egʰʷ-ei} `he is sober' have a labial \ipa{pʰ} from *\ipa{gʰʷ}, though \ipa{tʰ} would have been expected in the third person. 

Yet, this does not count as analogical levelling from the first person to the third, as the third person \textbf{plural} also has a non-front vowel \grec{νήφουσι} \ipa{nḗpʰousi}  $\leftarrow $ *\ipa{ne-egʰʷ-onti} `they are sober', and some tenses, such as the future (\grec{νήψει} \ipa{nḗpsei} $\leftarrow $ *\ipa{ne-egʰʷ-s-ei} `he will be sober'), have consonant-initial endings where the labial is expected in all forms by applying the regular sound laws.

A genuine case of analogical levelling applying from a first or second person to the third person singular should be such that the analogical change in question cannot be accounted for by analogy from another third person form, be it from non-singular forms in the same paradigm or from other tenses. 

Given the extreme rarity of first to third person analogical changes,   clear examples of this type from any language family would be relevant to our general understanding of language change.

 Siouan languages  present one uncontroversial example  of analogy from the first person to the second and third,  involving the verb `to think'. This paper presents a detailed account of this analogical change and provides possible explanations as to why specifically this verb underwent analogy in this direction. Most of the work on comparative Siouan is unpublished (though available from the Siouan Archive, which is freely distributed to all interested scholars), so that this paper will first present the accepted knowledge on Siouan reconstruction, mainly based on the model in the Comparative Siouan Dictionary (\citealt{csd2006}). 

 The first section presents the correspondences and reconstructions for the  consonants *\ipa{y}, *\ipa{r} and *\ipa{R}, which are relevant for understanding the case of analogy discussed in this paper. The following section discusses the main clusters containing these consonants. Then, we describe the verb paradigms of *\ipa{r} stem verbs, and show that the paradigm of the verb  \ipa{ebðé} `I think' in Omaha, which originally belonged to a different conjugation type, was remodelled following the *\ipa{r} stem verb conjugation on the basis of the first person: the third  person is renewed, while the first person is inherited. Finally, we discuss a possible explanation for the directionality of analogy in the case of this verb.
 

 
 

\section{Reflexes of simple *r, *R and *y in Dakotan and Dhegiha}
There is evidence in Mississippi valley Siouan languages of at least three distinct proto-phonemes *\ipa{r}, *\ipa{R} and *\ipa{y} before oral vowels. These phonemes merge as \ipa{r} in Chiwere and Winnebago, but are kept partially distinct in Dakotan and Dhegiha, according to the   correspondences in   Table \ref{tab:basic} (correspondences before nasal vowels are not indicated). 

These correspondences are valid word-initially or between vowels, except those involving the clusters *\ipa{ky--} and \ipa{wy--} (see below). In Kansa and Osage, the outcome of *\ipa{R} are palatalized as \ipa{ǰ} and \ipa{c} respectively before \ipa{e} (merging with *\ipa{ht}).

\begin{table}[h]
\caption{Basic correspondences of proto-Siouan *r, *R and *y in Dakotan and Dhegiha languages} \label{tab:basic} \centering
\begin{tabular}{l|ll|lllll}
\toprule
  & 	Lakhota  & 	Yankton/Sisseton & 	 Omaha  & 	 Kansa  & 	 Osage  & 	 Quapaw  &	\\	
  \midrule
\ipa{*y} & 	\ipa{čh} & 	\ipa{čh} & 	\ipa{ž} & 	\ipa{ž} & 	\ipa{ž} & 	\ipa{ž} & 	\\	
\ipa{*r} & 	\ipa{y} & 	\ipa{y} & 	\ipa{ð} & 	\ipa{y} & 	\ipa{ð} & 	\ipa{d} & 	\\	
\ipa{*R } & 	\ipa{l} & 	\ipa{d} & 	\ipa{n} & 	\ipa{d / ǰ} & 	\ipa{t / c} & 	\ipa{t} & 	\\	
\ipa{*r / i\_} & 	\ipa{čh} & 	\ipa{čh} & 	\ipa{ð} & 	\ipa{y} & 	\ipa{ð} & 	\ipa{d} & 	\\	
\midrule
\ipa{*kr} & 	\ipa{gl} & 	\ipa{hd} & 	\ipa{gð} & 	\ipa{l} & 	\ipa{l} & 	\ipa{kd} & 	\\		
\ipa{*xr} & 	\ipa{ȟl} & 	\ipa{hd} & 	\ipa{xð} & 	\ipa{xl} & 	? & 	? & 	\\		
\ipa{*wr} & 	\ipa{bl} & 	\ipa{md} & 	\ipa{bð} & 	\ipa{bl} & 	\ipa{br} & 	\ipa{bd} & 	\\		
\midrule
\ipa{*ky} & 	\ipa{čh / kč} & 	\ipa{čh / kč} & 	\ipa{gð} & 	\ipa{l} & 	\ipa{l} & 	\ipa{kd} & 	\\	
\ipa{*xy} & 	\ipa{ȟč} & 	\ipa{ȟč} & 	\ipa{xð} & 	\ipa{xl} & 	? & 	? & 	\\	
\ipa{*wy} & 	\ipa{čh / pč} & 	\ipa{čh / pč} & 	\ipa{bð} & 	? & 	\ipa{br} & 	? & 	\\	
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Examples of the correspondences of proto-Siouan *r, *R and *y in Dakotan and Dhegiha languages} \label{tab:basic.ex} \centering
 \resizebox{\columnwidth}{!}{
\begin{tabular}{ll|l|lll|ll}
\toprule
&proto-Siouan &Lakhota    & Omaha & Kansa & Osage   \\
	\midrule	
\ipa{*y} &	\ipa{*yo\_phe} &	\ipa{čho\_pȟÁ} &		\ipa{žúhe} &	\ipa{žophé} &	\ipa{žó\_pše} &		wade &	\\
&	\ipa{*yá•pE} &	\ipa{čhápa} &		\ipa{žábe} &	\ipa{žábe} &	\ipa{žápe} &		beaver &	\\
&	\ipa{*yá•že} &	\ipa{čhažé} &		\ipa{žáže} &	\ipa{žá•že} &	\ipa{žáže} &		name &	\\
&	\ipa{*yeká} &	\ipa{čhečá} &		\ipa{žegá} &	\ipa{žegá} &	\ipa{žeká} &		thigh, leg &	\\
&	\ipa{*yá•phe} &	\ipa{čha\_pȟÁ} &		\ipa{žáhe} &	\ipa{žáphe} &	\ipa{} &		stab &	\\
&	\ipa{*e/aya-ʔį} &	\ipa{ečhíŋ, ečháŋmi} &		\ipa{éžǫmį} &	\ipa{ažį́, ažamį} &	\ipa{ažį́, ážamį} &		think &	\\
	\midrule	
\ipa{*r-} &	\ipa{*ra-} &	\ipa{ya-} &		\ipa{ða-} &	\ipa{ya-} &	\ipa{ða-} &		by mouth &	\\	
&	\ipa{*ru-} &	\ipa{yu-} &		\ipa{ði-} &	\ipa{yü-} &	\ipa{ðu-} &		by hand &	\\	
&	\ipa{*xurá} &	\ipa{ȟuyá} &		\ipa{xiðá} &	\ipa{xüyá} &	\ipa{xúða} &		eagle &	\\	
&	\ipa{*rÉ} &	\ipa{yA´} &		\ipa{ðé} &	\ipa{yé} &	\ipa{(a)ðée} &		go &	\\	
	\midrule

 \ipa{*R} &	\ipa{*Ró•te} &	\ipa{loté} &		\ipa{núde} &	\ipa{dó•ǰe} &	\ipa{tóoce} &		throat &	\\
&	\ipa{*(i-)Rekší} &	\ipa{lekší} &		\ipa{inégi} &	\ipa{iǰégi} &	\ipa{įcéki} &		MB &	\\
&	\ipa{*Rá} &	\ipa{lá} &		\ipa{wana+} &	\ipa{da} &	\ipa{tá} &		ask &	\\
&	\ipa{*Ré•že} &	\ipa{léžA} &		\ipa{néže} &	\ipa{ǰéže} &	\ipa{céže} &		urine &	\\
&	\ipa{*Ré} &	\ipa{o\_lé} &		\ipa{u\_né} &	\ipa{o\_ǰé} &	\ipa{o\_cé} &		hunt, look for &	\\

\bottomrule
\end{tabular}}
\end{table}
Table \ref{tab:basic.ex} illustrates the first three correspondences; in this table and other Tables below, only Lakhota, Omaha, Kansa and Osage are presented. The cognates sets are from \citet{csd2006}, but the Lakhota, Osage  data were rechecked against     \citet{ullrich08} and \citet{quintero10osage} respectively and adapted to the spellings in these dictionaries. The Omaha data was rechecked against the Dorsey text corpus (\citealt{dorsey1890cegiha}; the forms not found in the corpus are indicated by the symbol +, and are taken directly from the comparative dictionary).
 

In some possessed nouns (especially body parts), we find a special correspondence with \ipa{čh} in Dakotan against \ipa{ð} : \ipa{y} : \ipa{ð} : \ipa{d} in Dhegiha, as in Table \ref{tab:ir}. This correspondence is interpreted as the effect of paradigm levelling, due to a sound change *\ipa{r}-- $\rightarrow$ \ipa{čh} after the third person possessive prefix \ipa{i--}, even before nasal vowels (see \citealt{rankin05quapaw}, \citealt{csd2006}). It is never attested in verbs, even those with a locative \ipa {í--} prefix; for instance *\ipa{íre} regularly yields  \ipa{íðe} 'see, find' in Omaha and \ipa{iyé\_ya} ‘find' in Lakota (not *\ipa{ičhé}).

\begin{table}[h]
\caption{Proto-Siouan *r after the possessive *i-- prefix} \label{tab:ir} \centering
 \resizebox{\columnwidth}{!}{
\begin{tabular}{ll|l|lll|ll}
\toprule
&proto-Siouan &Lakhota  &   Omaha & Kansa & Osage   \\
	\midrule	
&	\ipa{*i-ré•ži} &	\ipa{čheží} &		\ipa{teðéze} &	\ipa{yéze} &	\ipa{ðéeze} &		tongue &	\\
&	\ipa{*i-rá•ɣu} &	\ipa{čhaǧú} &		\ipa{ðǫ́xį+} &	\ipa{yáxü} &	\ipa{ðǫɣu} &		lung &	\\
&	\ipa{*i-ráke} &	\ipa{čhaká} &		\ipa{} &	\ipa{ho yáge} &	\ipa{} &		palate / gills &	\\
&	\ipa{*i-rą́•h-ka} &	\ipa{čhaŋkȟáhu} &		\ipa{nǫ́kka} &	\ipa{ną́kka} &	\ipa{ną́hka} &		spine &	\\
\bottomrule
\end{tabular}}
\end{table}




Before nasal vowels, *\ipa{r} and *\ipa{R} merge as \ipa{n} in Dakotan and Dhegiha languages in nearly all contexts (see  \citealt{michaud-jacques12nasalite}), but this topic will not be discussed in the present paper, as it has no incidence on the issue at hand.

In addition, there is a residue of words showing irregular correspondences, such as   `mosquito' (Lakota \ipa{čhaphų́ka}, Omaha \ipa{náhǫga}, Kansa \ipa{yáphąįge}, Osage \ipa{yáphąįge}, Quapaw \ipa{daphąke} ), but these case are most likely Wanderwörter and do not reflect the correspondences of inherited words. Therefore, they are not taken into account here.

\section{Reflexes of clusters}
There were some clusters with *\ipa{r} or *\ipa{y} as second element in proto-Mississippi Valley Siouan (the contrast between  *\ipa{r} and *\ipa{R} was neutralized). Examples of *\ipa{wr--}, \ipa{kr--},  \ipa{sr--}, *\ipa{šr--} and *\ipa{xr--} are plentiful, as can be illustrated by the data in Table \ref{tab:kr} (many other additional examples could be added).


 \begin{table}[h]
\caption{Proto-Siouan *\ipa{kr--}, *\ipa{xr--} and *\ipa{wr--}} \label{tab:kr} \centering
 \resizebox{\columnwidth}{!}{
\begin{tabular}{ll|l|lll|ll}
\toprule
&proto-Siouan &Lakhota    & Omaha & Kansa & Osage     \\
	\midrule	
*\ipa{kr--}&	\ipa{*kré•pE} &	\ipa{glépA} &		\ipa{gðébe} &	\ipa{lébe} &	\ipa{} &		vomit &	\\
&	\ipa{*krézE} &	\ipa{glézA} &		\ipa{gðéže} &	\ipa{léze} &	\ipa{léze} &		spotted &	\\
&	\ipa{*krá} &	\ipa{gla-} &		\ipa{ðigða+} &	\ipa{layá} &	\ipa{} &		untie &	\\
 	\midrule	
*\ipa{xr--}&	\ipa{*xroke} &	\ipa{ȟlókA} &		\ipa{xðúge} &	\ipa{xlóge} &	\ipa{} &		hole, nostrils &	\\
&	\ipa{*xro} &	\ipa{ȟló} &		\ipa{xðúde} &	\ipa{} &	\ipa{} &		snort &	\\
	\midrule	
*\ipa{wr--}	&	\ipa{*awró} &	\ipa{abló} &		\ipa{} &	\ipa{abló} &	\ipa{ábro} &		shoulder &	\\
&	\ipa{*hąwre} &	\ipa{haŋblé} &		\ipa{hǫ́bðe} &	\ipa{hąblé} &	\ipa{hǫ́bre} &		dream &	\\
&	\ipa{wráska} &	\ipa{blaská} &		\ipa{bðáska} &	\ipa{bláska} &	\ipa{bráaska} &		flat &	\\
&	\ipa{*wréh-(ka)} &	\ipa{blečá-} &		\ipa{bðékka} &	\ipa{blékka} &	\ipa{bréhka} &		thin &	\\
	\bottomrule
\end{tabular}}
\end{table}

Examples of clusters with \ipa{y} as a second element are much more restricted, and   Table  \ref{tab:ky} contains all known examples (from \citealt{koontz83rstem}  and \citealt{csd2006}).

 \begin{table}[h]
\caption{Proto-Siouan *\ipa{ky--}, *\ipa{xy--} and *\ipa{wy--}} \label{tab:ky} \centering
 \resizebox{\columnwidth}{!}{
\begin{tabular}{ll|l|lll|ll}
\toprule
&proto-Siouan &Lakhota    & Omaha & Kansa & Osage     \\
	\midrule	
*\ipa{ky--}&	\ipa{kyé•wrą} &	\ipa{wikčémna} &		\ipa{gðéba} &	\ipa{léblą} &	\ipa{lébrą} &		ten &	\\
&	\ipa{kyetą́} &	\ipa{čhetáŋ} &		\ipa{gðedǫ́} &	\ipa{ledą́} &	\ipa{letǫ́} &		hawk &	\\
&	\ipa{*kyąšká} &	\ipa{čhaŋšká} &		\ipa{gðąšká+} &	\ipa{} &	\ipa{} &		 &	\\
&	\ipa{*rukyą} &	\ipa{i\_yúkčaŋ} &		\ipa{wa-ðígðǫ} &	\ipa{í\_yülą } &	\ipa{í\_ðilą} &		know &	\\
	\midrule	
*\ipa{xy--}	&	\ipa{*xyá} &	\ipa{ȟčá / waȟčá} &		\ipa{waxðá} &	\ipa{xlá} &	\ipa{} &		bloom, flower &	\\
	\midrule	
*\ipa{wy--}&	\ipa{*e-w-ye} &	\ipa{epčÁ} &		\ipa{ebðé} &	\ipa{} &	\ipa{ébre}  &	I think &	\\	
&	\ipa{*rąwyE} &	\ipa{napčÁ}  &	\ipa{} &	\ipa{} &	\ipa{wa-nǫ́bre} &		eat, swallow &	\\	
&	\ipa{*w-yi--} &	\ipa{čhi-}  &	\ipa{wi--} &	\ipa{} &	 	\ipa{} &		1$\rightarrow$ 2 prefix& 	\\	
\bottomrule
\end{tabular}}
\end{table}

In Dhegiha, clusters in *\ipa{Cy--} merge with their *\ipa{Cr--} counterparts, except perhaps *	\ipa{wy--} if the 1$\rightarrow$ 2 prefix \ipa{wi--} is the regular outcome of \ipa{*w-yi--}, in which case *\ipa{wy--} perhaps present a different treatment word-initially.

In Lakhota, the cluster *\ipa{y--} element in *\ipa{Cy--} cluster changes to \ipa{č} word-internally, and the previous segment is preserved (and undergoes fortition, in the case of *\ipa{wy--}). Word-initially, *\ipa{wy--} and *\ipa{ky--} apparently merge with *\ipa{y} as 	\ipa{čh--}.\footnote{The 1$\rightarrow$ 2 prefix does not always appear word-initially, but we can safely assume that analogical pressure would have removed the expected \ipa{čh--} / \ipa{pč--} allomorphy.}
%wyi > chi- (wi- in O

 
%\citet{rankin02ofo}
\section{Verbal paradigms}
 
The paradigm of *\ipa{r--} initial verb stems (including verbs with the instrumental prefixes *\ipa{ra--} `by mouth' and *\ipa{ru--} `by hand') constitute a distinct conjugation class in Mississippi Valley Siouan languages. The paradigms in these languages are provided in Table \ref{tab:go} (see \citealt{koontz83rstem, koontz90syncopating}). 

The first and second person forms result from the first singular *\ipa{w--} and second person *\ipa{y--} prefixes respectively which occur in the so-called `syncopating paradigms'.\footnote{The non-syncopating  regular active and stative paradigms, whose correspondences are quite complex between Siouan languages, will not be discussed here, as they are not relevant to the topic at hand.} In all languages except Lakota, the *\ipa{y--} prefix becomes \ipa{š} in this paradigm by obstruentization.\footnote{The absence of \ipa{š} in Lakota might be secondary; in Omaha, second person forms of  *\ipa{r--} stem verbs have variants in \ipa{hn} and \ipa{n} instead of  \ipa{šn} in the text corpus.}. That *\ipa{y--}, rather than *\ipa{š--}, should be reconstructed here is shown by other syncopating paradigms such as that of the glottal stop stems; for instance, the second person form of \ipa{ʔǫ́} `be, do' in all Dhegiha languages is \ipa{žǫ́}  `you do, you are', which can only be from *\ipa{y-ã} $\leftarrow$ *\ipa{y-ʔã}. 

%of the first person *\ipa{wa--} and second person  \ipa{ya--} prefixes.\footnote{The reconstruction of the second person prefix is a complex problem: Mississippi Valley Siouan languages point to *\ipa{ra--}, while Biloxi and Ofo suggest a reconstruction *\ipa{ya--} instead. This issue is in any case irrelevant to the present  topic, as we focus exclusively on MVS. }


\begin{table}[h]
\caption{Paradigm of the verb *\ipa{re} `go' in Siouan languages} \label{tab:go}
 \resizebox{\columnwidth}{!}{
\begin{tabular}{ll|l|ll|llll|ll}
\toprule
 &	 &	Lakhota &	Winnebago &	Chiwere &	Omaha &	Osage &	Kansa &	Quapaw &	Ofo &	\\	
  \midrule
1sg.A &	\ipa{*wre•} &	\ipa{blÁ} &	\ipa{tée} &	\ipa{hajé} &	\ipa{bðe} &	\ipa{brée} &	\ipa{bne} &	\ipa{bde} &	\ipa{até•kna} &	\\	
2sg.A &	\ipa{*šre•} <\ipa{*yre•}  &	\ipa{lÁ} &	\ipa{šeré} &	\ipa{slé} &	\ipa{šne} &	\ipa{šcée} &	\ipa{hne} &	\ipa{tte} &	\ipa{šté•kna} &	\\	
base &	\ipa{*re•} &	\ipa{yÁ} &	\ipa{rée} &	\ipa{lé} &	\ipa{ðé} &	\ipa{aðée} &	\ipa{yé} &	\ipa{dé} &	\ipa{té•kna} &	\\	
\bottomrule			

\end{tabular}}
\end{table} 
 
 In Omaha and other Dhegiha languages, the verb \ipa{ebðé}, \ipa{éðe} `think' follows the same conjugation as *\ipa{r} stem verbs, as shown in Table \ref{tab:think}.\footnote{This verb has a discontinuous stem  \ipa{e\_ðe}. The element \ipa{e--} is probably originally a demonstrative related to Omaha \ipa{é}  that'.} However, its Lakota cognate  \ipa{epčÁ} `I think', a defective verb attested only in the first singular, belongs to the set of forms with *\ipa{wy--} in proto-Siouan. The only other example of the cluster *\ipa{wy--} is Lakota \ipa{napčÁ}  `swallow'	with Osage \ipa{wa-nǫ́bre} `eat (it)', and these two forms can be reconstructed as proto-Mississippi Valley Siouan *\ipa{rãwye}.\footnote{The \ipa{wa--} prefix in Osage is the antipassive. This word is apparently not attested in other Dhegiha languages.} 
 
 Since the cluster \ipa{pč} is extremely rare in Lakota, the dearth of cognates is not surprising. Moreover, the parallelism with the group *\ipa{ky--}, which merges with *\ipa{kr--} in Dhegiha but remains distinct as \ipa{kč} in Lakota, confirms that *\ipa{wy--} is indeed the correct reconstruction for the correspondence \ipa{pč} in Lakhota to \ipa{bð} in Omaha.
 
 \begin{table}[h]
\caption{Paradigm of the verb *\ipa{e\_ye} `think' in Siouan languages}  \label{tab:think} \centering
\begin{tabular}{ll|l|ll|llll|ll}
\toprule
 &	 &	Lakhota &	 	Omaha &	Osage 	& Expected Omaha\\	
 \midrule
1sg.A &	*\ipa{e-w-ye} &	 \ipa{epčÁ} &   \ipa{ebðé} &\ipa{ébre} &\ipa{ebðe}\\
2sg.A &*\ipa{e-y-ye}&  &  \ipa{ešné} &? &???& \\
base &	*\ipa{e-ye} &	  &\ipa{eðé} &\ipa{éðe} & *\ipa{eže}\grise{}\\
\bottomrule			
\end{tabular}
\end{table} 

 Thus, in the Dhegiha paradigm for the verb \ipa{ebðé} `I think'     the first person should  be identical to that of an \ipa{r--} stem, since *\ipa{wy} and *\ipa{wr} merge as \ipa{bð} in Omaha and \ipa{br} in Osage according to the correspondences shown in Table \ref{tab:ky}. However,  the third person should not have the same reflexes in Dhegiha:  *\ipa{e-ye} should have yielded *\ipa{eže} instead of the attested form \ipa{eðé} in Dhegiha. The third person \ipa{eðé}  `he thinks' is the expected  form of an *\ipa{r} stem verb, not that of a *\ipa{y} stem verb.
 
 
 For the second person, it is unclear what the outcome of *\ipa{e-y-ye} or *\ipa{e-š-ye} would have been; while  *\ipa{eže}  is perhaps the most likely outcome, we cannot exclude the possibility that the second person form  \ipa{ešné} is the regular phonetic reflex from  *\ipa{e-y-ye} (since *\ipa{y} otherwise merges with *\ipa{r} when occurring as second element of clusters), in the absence of other examples of *\ipa{--yy--} in Siouan.

The only available explanation for the irregular \ipa{ð} in the form \ipa{eðé} `he thinks' is analogical levelling based on the model of *\ipa{r} stem paradigms, taking the first person as the pivot form from which analogy spread. It is impossible to argue that \ipa{ð} is the regular outcome of *\ipa{y} in Dhegiha in intervocalic position, since clear examples of \ipa{ž} from *\ipa{y} are attested word-internally (see Table \ref{tab:basic}).

The verb *\ipa{e\_ye} `think' in Siouan was the only syncopating *\ipa{y--} initial stem, and had a unique conjugation. In Lakota it lost the second and third person forms (it is the only defective verb of this type), while in Dhegiha the third and perhaps the second person forms were renewed on the basis of the first person singular.


\section{The directionality of analogy}

Mississippi Valley Siouan languages attest  a very rare type of analogy, from first person to third person, which as we saw before is almost non-attested in Indo-European languages. Unlike Attic Greek, in Omaha the third person plural  and TAM morphology are marked by suffixes, so that no other form of the paradigm could have served as a basis for the \ipa{ð} in \ipa{éðe} `he thinks'.

An explanation for the preservation of the first person form \ipa{epčÁ} `I think' in Lakhota (and the loss of the rest of the paradigm) and the directionality of analogy in the paradigm of \ipa{ebðé}, \ipa{éðe} `think' in Dhegiha can however be provided by the relative frequency of the first person singular for verbs of cognition such as   `think' and `know'.

 %(unfortunately, no example of  \ipa{epčÁ} `I think' could be found in \citealt{deloria32dakota} to confirm whether this use is attested).\citet{dorsey1890cegiha}
The Omaha corpus (\citealt{dorsey1890cegiha}) was conveniently retranscribed and corrected by Robert Rankin (\citealt{rankin08dhegiha}), and is therefore fully searchable. As shown in Table \ref{tab:cognition}, the most common  verbs of cognition meaning `think' or `know' in Omaha occur most often in the first person singular than in the third person by a considerable margin. This strong tendency is verified in other languages too; Table \ref{tab:cognition2} presents the token counts of the singular forms of the verb `to think' in French, English and Chinese (in Google).

 

 \begin{table}[h]
\caption{Number of attestations of verb forms in the Omaha corpus (`think', `suspect' and `know')}  \label{tab:cognition} \centering
\begin{tabular}{l|ll|ll|lll}
\toprule
Person	&think &	&	think, suspect &	&	know &	&	\\
 \midrule
1 & 	\ipa{ebðé(gǫ)}  & 	172 & 	\ipa{éžǫmį} & 	7 & 	\ipa{iðáppahǫ} & 	42 & 	\\
2 & 	\ipa{ehné(gǫ)},  & 	114 & 	 & 	0 & 	\ipa{íšpahǫ} & 	28 & 	\\
& \ipa{ené(gǫ)} or \ipa{ešné(gǫ)} &&&&\\
3 & 	\ipa{eðé(gǫ)} & 	22 & 	 & 	0 & 	\ipa{íbahǫ} & 	29 & 	\\
	\bottomrule			
\end{tabular}
\end{table} 


 \begin{table}[h]
\caption{Paradigm of the verb `to think' in several languages (in millions of tokens)}  \label{tab:cognition2} \centering
\begin{tabular}{l|ll|ll|lll}
\toprule
Person	&French & & English && Chinese && \\
 \midrule
1 & \ipa{je pense} & 96.9 & \ipa{I think}  & 1290 & \zh{我想} &40.9 \\
2 & \ipa{tu penses} & 5.6 & \ipa{you think}  &777& \zh{你想} &23.1 \\
3 &  \ipa{il/elle pense} & 4.4 & \ipa{He/she thinks}  &61.7& \zh{他/她想} &14.1 \\
	\bottomrule			
\end{tabular}
\end{table} 

%nudoⁿhoⁿga akʰá tʼeóⁿtha-bázhi ai thóⁿshti, á-biamá

%óⁿkʰazhi, nudóⁿhoⁿga akʰá é wakʰa-bázhi ebthégoⁿ, á-bi-amá

The verbs \ipa{ebðégǫ} and  	\ipa{éžǫmį} in Omaha, as well as   \ipa{epčÁ} `I think' in Lakota can only be used with a complement clause (in Omaha, the verb \ipa{síðe} `think of' is used when the P is not a complement clause).

The first person form of the verb   \ipa{ebðégǫ}  `to think'  can be used  in dialogue to present one's point of view, either as  opposed to someone else's (as in  example   \ref{ex:ebdhegon}; sentence \ref{ex:no.ebdhegon} is repeated a few lines below  with the verb \ipa{ebðégǫ} `I think' added to it), or as a more general assertion, as in \ref{ex:ebdhegon2}.

% is particularly common in the  
 
\begin{exe}
 \ex \label{ex:ebdhegon} 
 \gll ``\ipa{ǫ́kʰaži,}  	\ipa{nudǫ́hǫga}  	\ipa{akʰá}  	\ipa{é}  	\ipa{wakʰa-báži}  	\ipa{ebðégǫ}"  	\ipa{á-bi-amá}\\
no leader \textsc{prox:sg} that mean-\textsc{neg} \textsc{1sg}:think say-\textsc{prox-evd}\\
%\end{exe}
%\begin{exe}
 \ex \label{ex:no.ebdhegon}
 \gll ``\ipa{ǫ́kʰaži,}  \ipa{ha}	 \ipa{nudǫ́hǫga}  	\ipa{akʰá}  	\ipa{é}  	\ipa{wakʰa-í}"   	\ipa{á-bi-amá}  \\
no \textsc{interj} leader \textsc{prox:sg} that mean-\textsc{prox}  say-\textsc{prox-evd} \\
\glt ``No, the leader did not mean that, I think," said the former. ``Yes, the leader did mean that," said the latter. (177:19-178.1)
\end{exe}

%Alternatively, they are also used to introduce  sentences that is are necessarily strictly the speaker's point of view, but belong  to the commonly accepted knowledge of the culture, as in \ref{ex:ebdhegon2}.

\begin{exe}
 \ex \label{ex:ebdhegon2}
 \gll 
\ipa{nį́kašįga}  	\ipa{ðižúbaži}  	\ipa{kíǰi,}  	\ipa{égǫwéʼǫ}  	\ipa{gǫðawáðe}  	\ipa{ebðégǫ.}  \\
man injure when do.this.way desirable \textsc{1sg}:think \\
\glt I think that when one man injures another, it is desirable to repay him. ( 438,  12)
  \end{exe}

In all these examples, the verb  \ipa{ebðégǫ}  `I think' is semantically closer to a discourse marker than to an actual predicate, which explains its higher frequency. 



Outside of Siouan, no examples of analogy based on the first person for verbs of cognition have been described. Likewise, it is unclear whether we find defective verbs only attested in the first person like \ipa{epčÁ} `I think'  in any other family. The closest case would be English \ipa{methinks}, which, although not verbal in the strict sense, is only possible in the first person (*\ipa{youthinks}, *\ipa{himthinks} are not attested) and is very close semantically to the verb form `I think'.

It is possible that similar cases of analogy from the first person to the third and second wait to be discovered in other language families; the search for such cases should be focused on verbs of cognition such as `think',  `know', `believe' or `suspect', but possibly also on any class of verbs which are significantly more frequent in the first person than in the third.
   
%níⁿkashiⁿga thizhúbazhi ki´ǰi, égoⁿwéʼoⁿ góⁿthawáthe ebthégoⁿ.

 
%This example does not disprove the validity of Watkins' law; rather, it can help to refine it. 



%generic forms are easily interpreted as being first person and vice versa. The semantic proximity between first person and generic in the case of these verbs is due to two opposite reasons. First, using a generic rather than a first person is a way for the speaker to avoid directly asserting his own point of a view. Second, a confusion between first person and generic can occur when the speaker equates what he thinks or believes with `common sense' or common knowledge.
%ehé, níⁿkashiⁿga thizhúbazhi ki´ǰi, égoⁿwéʼoⁿ góⁿthawáthe ebthégoⁿ.
%said (one). / Oho! / I said, / man / injures one / when, / to do so in return / desirable / I think. /
% I think that when one man injures another, it is desirable to repays him (438, line 12.)

%
%A example of semantic confusion between first person and generic with cognition verbs can be observed  in Japhug (Sino-Tibetan, \citealt[342]{jacques04these}, \citealt{jacques15generic}). In Japhug, the highly irregular negative generic form \ipa{mɤ-xsi} `one does not know' of the verb \ipa{sɯz} `know' is used in contexts where a first person singular (not first person plural) is  implied, and a generic interpreted should be excluded, as in examples \ref{ex:mAxsi} and \ref{ex:mAxsi2}.
%
%\begin{exe}
% \ex \label{ex:mAxsi}
% \gll
%\ipa{ma-tɯ-nɯqaɟy}  	\ipa{ma}  	\ipa{ɕu}  	\ipa{tɯ-ŋu}  	\ipa{mɤ-xsi}  	\ipa{ri} \ipa{kɯ-qarŋe}  	\ipa{thamtɕɤt}  	\ipa{nɯ}  	\ipa{rɟɤlpu}  	\ipa{ɣɯ}  	\ipa{χsɤrɲa}  	\ipa{ŋu,}  	
% \\
%\textsc{neg:imp}-2-fish because who 2-be:\textsc{fact} \textsc{neg-genr}:know but nmlz:S/A-be.yellow all \textsc{dem} king \textsc{gen} gold.fish be:\textsc{fact}\\
%\glt Don't fish; although I don't know who you are, the yellow ones are the King's gold fishes (ie, only the king can fish them, so whoever you are, you are not allowed). (Gesar, 368)
%\end{exe}
%\begin{exe}
% \ex \label{ex:mAxsi2}
% \gll
%\ipa{mɤ-xsi}  	\ipa{ko,}  	\ipa{nɯra}  	\ipa{ɲɤ-nɯ-jmɯt-a}  \\
%\textsc{neg-genr}:know \textsc{sfp} \textsc{dem:pl} \textsc{evd-auto}-forget-\textsc{1pl} \\
%\glt I don't know, I forgot those things. (Conversation, answer to the question: `How did you talk to your children when they were small babies?')
%\end{exe}
%
%Note that generic verbs forms are never used for first person singular with other verbs. Additionally, some languages have adverbs meaning `who knows' or `I don't know' (for instance Nepali \ipa{kunni}),  whose use is comparable to the Japhug generic form \ipa{mɤxsi}. 

\section{Conclusion}
In view of the analogy in the verb `to think' in Omaha, and the cases of defective verbs restricted to the first person, Watkins's law can be reformulated as \ref{ex:watkins2}.

\begin{exe}
\ex  \label{ex:watkins2}
\glt Analogy takes place from the most frequent form to the least frequent one (in discourse). 
\glt When the most frequent form is third person singular, it can be reinterpreted as a zero-marked form
\glt  When the most frequent form is  is first person (in the case of verbs of cognition), the other forms are remodelled by proportional analogy, by applying attested rules of correspondence between first person and other persons in a more regular paradigm.
\end{exe}

Therefore, hypotheses in historical linguistics involving analogy from the first to the  third person can be allowed in cases where it can be demonstrated, on the basis of corpus study, that the first person is more frequent than the third.
 
 
\bibliographystyle{linquiry2}
\bibliography{bibliogj}

 \tableofcontents
\end{document}