\documentclass[oldfontcommands,oneside,a4paper,11pt]{article} 
\usepackage{fontspec}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xltxtra} 
\usepackage{longtable}
\usepackage{polyglossia} 
\usepackage[table]{xcolor}
\usepackage{gb4e} 
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref} 
\hypersetup{bookmarks=false,bookmarksnumbered,bookmarksopenlevel=5,bookmarksdepth=5,xetex,colorlinks=true,linkcolor=blue,citecolor=blue}
\usepackage[all]{hypcap}
\usepackage{memhfixc}
\usepackage{lscape}

\bibpunct[: ]{(}{)}{,}{a}{}{,}

%\setmainfont[Mapping=tex-text,Numbers=OldStyle,Ligatures=Common]{Charis SIL} 
\newfontfamily\phon[Mapping=tex-text,Ligatures=Common,Scale=MatchLowercase,FakeSlant=0.3]{Charis SIL} 
\newcommand{\ipa}[1]{{\phon \mbox{#1}}} %API tjs en italique
\newcommand{\ipab}[1]{{\scriptsize \phon#1}} 

\newcommand{\grise}[1]{\cellcolor{lightgray}\textbf{#1}}
\newfontfamily\cn[Mapping=tex-text,Ligatures=Common,Scale=MatchUppercase]{MingLiU}%pour le chinois
\newcommand{\zh}[1]{{\cn #1}}
\newcommand{\refb}[1]{(\ref{#1})}


\XeTeXlinebreaklocale 'zh' %使用中文换行
\XeTeXlinebreakskip = 0pt plus 1pt %
 %CIRCG
 


\begin{document} 
\title{Evaluation of the ms. `A Web of Relations A grammar of rGyalrong Jiǎomùzú (Kyom-kyo) dialects'} 
\maketitle 
\sloppy

 This book is a very important contribution to the documentation of Rgyalrongic languages, and deserves to be published. A very positive aspect of this work is that it is based not only on elicitation, but also on narratives and conversations, and is based on a relatively rich corpus. It is the first grammar of a Rgyalrong language in English, and will therefore become a major reference in both linguistic typology and Sino-Tibetan comparative linguistics.  The dialect described here has never been studied in detail before, and is highly valuable for linguists and local people alike.
 
Since this draft will become a major reference work once it is published, it is preferable, before publication, to correct a certain number of infelicities.

An (easily correctable) problem with this ms. is that the references to the existing literature are very incomplete, and in some cases this has consequences on the the analysis. In the following, I will provide a list of references that must be included in the final version. These references that are either published or are MA/PhD dissertations, but in all cases they are either available from the publisher's website or the respective authors' pages on their university pages or on academia.edu, and there is thus no excuse for not using them.

Important recent work that should be referred to include Daudey's grammar of Pumi (\citealt{daudey14grammar}), Lai's work on Khroskyabs (\citealt{lai13affixale}), Jackson Sun's most recent articles on Tshobdun (\citet{sun12complementation}, \citet{sun14generic}, \citet{jackson07irrealis}, \citet{jackson14morpho}), Jacques's articles on Japhug (in particular \citet{jacques13tropative},\citet{jacques12incorp}, \citet{jacques13harmonization}, \citet{japhug14ideophones}, \citet{jacques14linking}, \citet{jacques14antipassive}), Gong Xun's article (\citealt{gongxun14agreement}) and Lin Youjing's most recent articles (\citealt{lin11direction},  \citealt{linyj12tone}).

Moreover, I strongly advise against using Chinese-based transcriptions for places and language names, especially since `Japhug' and `Tshobdun' are the names used in English by the specialists of these languages, and have been used in general references works (see for instance \citealt{jackson14morpho} or \citealt{jacques13harmonization}), and its is by these names that typologists know about these languages. Using `Chabao' and `Caodeng' instead will cause confusion for typologists using this book, and is offensive to linguists working on these languages. This issue really needs to be corrected for this book to be published, at least for these two languages.

It is somewhat contradictory to use both the Tibetan-based name`rGyalrong' and the Chinese transcription `Jiaomuzu'. I think that `Kyomkyo' (or Kyomukyo) would be better for several reasons. First, it would be more readable for readers unfamiliar with pinyin. Second, the Chinese \zh{脚木足} is actually read `Jiaomujiao' instead of `Jiaomuzu' locally. 

A highly cited typologist has recently suggested to use a simplified orthography for rGyalrong/Rgyalrong (Like for instance Gyalrong), as the ``rgy" cluster looks difficult for non-specialist, and may deter people from reading articles and books about `languages with an unpronounceable name". I leave it to the author to decide what to do with such a suggestion, but I thought that I should relay it here.


 \section{Corrections and additional references}
Many Kyomkyo verbal forms are not properly analyzed into morphemes. In particular, verbs are generally quoted in their infinitive forms but the infinitive prefixes \ipa{ka-- / kə--} are not separated from the verb stem by a hyphen (the same for derivational prefixes such as the causative, applicative etc). It is important to help readers to properly parse the forms.


\begin{itemize}
\item Indicate how `tusi' are called in Kyomkyo.
\item p7, Most specialists do not consider `Rgyalrong' to be a  language', but rather a group of at least four languages (Japhug, Zbu, Tshobdun and Situ) (see for instance \citealt[195]{jacques13harmonization})
\item p7, \citet{jacques.michaud11naish} propose a different subgrouping hypothesis, Burmo-Qiangic, which is also developed in \citet{jacques14esquisse}.
\item p9 The name `Lavrung' is now abandoned by specialists of this language (Lai Yunfan and his local collaborator Rig'dus Lhamo) in favour of the traditional term for this region, Khroskyabs (see for instance \citealt{delancey14second}, \citealt{jacques14inverse}). Please adopt it too (`Lavrung' is based on a misunderstanding on the part of Huang Bufan).
\item p10, the preservation of uvular in Zbu, Tshobdun and Japhug is really not the most important isogloss defining those languages (the verbal morphology provides many more useful criteria, in particular stem alternation, second person markers, the slot insertion of associated motion prefixes etc).
\item I am surprised to see that jiehun is borrowed with a voiced stop. In my experience Sichuan Chinese unaspirated stops/affricates are borrowed as unvoiced, not as voiced obstruents, at least word-initially.
\item pp 43-8 on tones, see \citet{linyj12tone}, who presents a very different analysis of the Situ tone system. See also \citet{jackson05yingao.zh}.
\item p86(+p88) ``rGyalrong is a head marking language" > Rgyalrong languages are strongly head marking (there is more than one Rgyalrong language)
\item pp 210, On the ergative: in Japhug, the ergative is really obligatory with overt third person agents, unlike in Kyomkyo. Also, the argative can be used with the sole argument of intransitive verbs, but in this case it implies the meaning `more than' (\citet[202]{jacques12demotion}); it is used in the comparative construction. I think that Rgyalrong languages differ considerably of in their use of the ergative, although it may be superficially similar.
\item pp 250-7, on expressives/ideophones, see also \citet{japhug14ideophones} and \citet{jackson14morpho}
\item pp 267, on conjunctions, see also \citet{jacques14linking}
\item p304, on the discussion of the verbal template, see also \citet[197-199]{jacques13harmonization}, \citet[12]{jacques14antipassive} and \citet{lai13affixale}. Does the verbal morphology of Kyomkyo present templatic features?
\item p311, ``I call this kind of verb root 'root 3'" > actually this is \citet{jackson00sidaba, jackson07irrealis}'s terminology, who deserves more explicit credit here, as he was the first to explicitly describe stem alternation in Rgyalrong languages.
\item Concerning the second person prefix in historical perspective, see the more recent work by \citet{jacques12agreement} and \citet{delancey14second}.
\item The tables in general, and p334-8 in particular, must be reorganized following the general system used for polypersonal paradigms (see \citet{gongxun14agreement}, \citet{jacques14inverse} for examples of how to do it)
\item These are not compounds. A more correct term would be `complex predicates' with light verbs. NB there are compound verbs in Rgyalrong languages, see \citet{jacques12incorp}, but this is not the same phenomenon).
\item pp 344+, on inverse marking see also \citet{gongxun14agreement}, \citet{jacques14inverse}, \citet{jacques14rtau} and \citet{lai14person}
\item ``In fact, a structure with third person observation marker na- as well as inverse marker wu- is not
grammatical:" this is strange, such phenomenon has never been mentioned previously. Maybe develop. Note that in Japhug the inverse prefix is \ipa{wɣɯ--}, not \ipa{wu--}.
\item pp 398 what the author refers to as `viewpoint markers' (\ipa{ʃi--} and \ipa{və--}) have been described as associated motion markers in the literature (\citealt{jacques13harmonization}). The author needs to take into consideration this article and evaluate to what extent the cognate markers \ipa{ɕɯ--} and \ipa{ɣɯ--} in Japhug differ or work in the same way as the \ipa{ʃi--} and \ipa{və--} prefixes in Kyomkyo. The author is free to keep the term `viewpoint markers', but he/she must address  the question whether a distinct analysis would not be preferable. Also, a comparison with the motion verb constructions, in the lines of \citet[203]{jacques13harmonization}, seems necessary (these are mentioned later in the ms, p541).
\item p398 \ipa{nənɟo məto-tə-laʔt-n me}: I am puzzled by \ipa{məto-tə-laʔt-n}, shouldn't we have \ipa{məto-tə-laʔt-w}, instead, since this is a transitive verb? And in any case, the \textsc{2sg} \ipa{--n} is never realized after a stop.
\item p406 I would advise against `observational', a more standard term is `testimonial' (\citealt{hill13hdug}) or `sensory' (\citealt{tournadre14evidentiality})
\item p410 Concerning `mirativity', see \citet{hill13hdug} and \citealt{tournadre14evidentiality}.
\item The chapter on verb derivations requires some rewriting:
\begin{itemize}
\item p433 On the applicative, see \citet{jacques13tropative} and \citet{jackson14morpho}. An interesting point, judging from the data in the ms, is that the applicative prefix in Kyomkyo seems much more widespread and productive than in Japhug. 
\item p433 delete the pair ``\ipa{kaməsem} listen, understand \ipa{karəkna} hear" from your table.
\item  p433 The term `anticausative' should appear here (see \citealt{jacques12demotion}). Although there is indeed non-volitionality in some cases, the crucial thing here is that the verbs on the second column are intransitive, while those on the first column are transitive. Also, please note that some anticausative verbs, when employed with a human/animal S, can be volitional (for instance in Japhug \ipa{mbɣaʁ} \zh{打滚}  can be used to refer to a dog that does it on purpose to play).
\item p437 see \citet{jacques10refl}
\item p447 see \citet{jacques07passif}, \citet{jacques12demotion} and \citet{jackson14morpho}; 
\item p454 The  term `antipassive' should appear here (see \citealt{jacques12demotion}, \citet{jacques14antipassive} and \citet{jackson14morpho}), and it should be described a a distinct voice.
\item Incorporation (\citealt{jacques12incorp}) is not described, though it certainly exists in Situ (see for instance the cognate of Japhug \ipa{nɯrɟɯrŋom} `covet riches', that is attested in Cogtse and Bragdbar). 
\item The tropative derivation certainly exists in Kyomkyo, but is not mentioned (see \citealt{jacques13tropative}, for instance the cognate of Japhug \ipa{nɤ-mpɕɤr} `to find beautiful').
\item The deexperiencer derivation is not described in this section, though some examples appear scattered in the ms, misanalyzed as `causatives' (see \citealt{jacques12demotion}, \citet{jackson14morpho})
\item Something should be said of denominal verbs in the `derivation' section (see \citealt{jacques14antipassive} and \citealt{jacques13tropative})
\end{itemize}
\item  p469-470 conditional should not be treated in the section on modality; rather, it should appear in chapter 8 on `complex sentences'). See \citet{jackson07irrealis} and \citet[295-303]{jacques14linking}
\item p470 Concerning indirect speech, see \citet[344-346]{jacques08zh} and \citet[310-311]{jacques14linking}
\item ``\ipa{stʃi}, which conveys a condescending sense of ‘be’": I am not sure `condescending' is a correct description of the difference between \ipa{stʃi} and \ipa{ŋos,}).
\item p534, on complement clauses, see \citet[3337-353]{jacques08zh}  and \citet{sun12complementation}.
\item pp543+, on purposive/manner clauses, see \citet{jacques14linking} for a comparison in Japhug. The term `converb' does not appear in this section; 
\end{itemize}
 

 
 
\section{Typos}
general remarks:
\begin{itemize}
\item  nominalization' should be glossed as \textsc{nmlz}, not \textsc{nom} (which is the standard gloss for `nominative').
\item I would suggest to include the Chinese characters for person and placenames in the body of the text rather than in footnote (it is distracting to the reader).
 \item Transcribing the verbs in their underlying form is useful, but sometimes it could also be helpful to the reader to have the surface realization, at least in the sections on morphology.
 \item for reduplication, use the Leipzig glossing rules convention \textasciitilde{}, thus for instance (from p445):
\begin{exe}
\ex  
\gll \ipa{ná-ŋa-sa\textasciitilde{}sat-ɲ} \\
\textsc{sens-recip}-\textsc{recip}\textasciitilde{}kill-\textsc{pl} \\
\glt  They are killing each other. 
\end{exe} 
\end{itemize} 
 specific:
\begin{enumerate}
\item p xi Ubniversity > university
\item  p157 ``A-myis Sgo-ldung dealt him a blow with the iron hammer. " > put the translation after the second line ``to-leʔt-jn"
\item p342 \ipa{to-najo-n}, the gloss 3/2-wait-1s should be changed to  \textsc{2:inv}-wait-\textsc{2sg}
\item p460 caustivity > causativity
\item p551 Why spelling A-myis rather  than A-myes?
\end{enumerate}  
  
  
\bibliographystyle{unified}
\bibliography{bibliogj}
\end{document}