\documentclass[oldfontcommands,oneside,a4paper,11pt]{article} 
\usepackage{fontspec}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xltxtra} 
\usepackage{longtable}
\usepackage{polyglossia} 
%\usepackage[table]{xcolor}
\usepackage{gb4e} 
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{float}
\usepackage{lineno}
\usepackage{textcomp}
\usepackage{hyperref} 
\hypersetup{bookmarks=false,bookmarksnumbered,bookmarksopenlevel=5,bookmarksdepth=5,xetex,colorlinks=true,linkcolor=blue,citecolor=blue}
\usepackage[all]{hypcap}
\usepackage{memhfixc}
\usepackage{lscape}
 

\setmainfont[Mapping=tex-text,Numbers=OldStyle,Ligatures=Common]{Charis SIL} \newfontfamily\phon[Mapping=tex-text,Ligatures=Common,Scale=MatchLowercase,FakeSlant=0.3]{Charis SIL} 
\newcommand{\ipa}[1]{{\phon #1}} %API tjs en italique
 
\newcommand{\grise}[1]{\cellcolor{lightgray}\textbf{#1}}
\newfontfamily\cn[Mapping=tex-text,Ligatures=Common,Scale=MatchUppercase]{MingLiU}%pour le chinois
\newcommand{\zh}[1]{{\cn #1}}


\XeTeXlinebreaklocale "zh" %使用中文换行
\XeTeXlinebreakskip = 0pt plus 1pt %
 %CIRCG
 


\begin{document} 

\title{Complementation\footnote{
} }
%\author{Guillaume Jacques}
\maketitle
\linenumbers

\section{Attention verbs}

\begin{enumerate}
\item see \ipa{vdɔ}

\begin{exe}
\ex
\gll [ŋa tə rka-sə] vdow \\
\\
\glt I saw that he was tired.
\end{exe}

\begin{exe}
\ex
\gll [ŋa ɲi rka-sə ci] vdow-rə \\
\\
\glt I see that you are tired. %(-rə = présent)
\end{exe}

\begin{exe}
\ex
\gll [ŋa   rka-sə ci]  mi-vdej-rə ŋɔ \\
\\
\glt Don't you see that I am tired. %(-rə = présent)
\end{exe}

\begin{exe}
\ex
\gll [ŋa   tsəgə χserba kə-ru-sə]  ma-vdej  ŋɔ \\
\\
\glt Didn't you see that I bought a new clothe? %(-rə = présent)
\end{exe}

\begin{exe}
\ex
\gll [ŋa   tsəgə χserba kə-ru-sə] de ə́-vdej    \\
\\
\glt Have you seen the new clothes that I bought? / Have you seen that I bought new clothes?%(-rə = présent)
\end{exe}
\begin{exe}
\ex
\gll [ɲi ʁə nə-praˠ-sə] ŋa χavdə demdi   ma-vdow  \\
\\
\glt  I just noticed that you had a haircut.
\end{exe}
\begin{exe}
\ex
\gll [ŋa   tsəgə χserba kə-ru-sə]-ʁa kí-scəqi?    \\
\\
\glt Have you had a look at the new clothes that I bought? /  %(-rə = présent)
\end{exe}


\item hear \ipa{ntɕʰi}
\begin{exe}
\ex
\gll ŋa [khakha-ɲu jə-sə ci-ŋkʰə] tə-ntɕʰu-ge, ɲi qəsji ɣə-kʰi ŋə-rə jə-rə    \\
\textsc{1sg} other-\textsc{pl:erg} \\
\glt  I hear from the other people that you will come tomorrow.
\end{exe}

\begin{exe}
\ex
\gll ŋa pʰjəsɔ məqʰi ɮde-sə ci-ŋkʰə ntɕʰu-rə    \\
\\
\glt  I hear that it rains outside.
\end{exe}

\begin{exe}
\ex
\gll ŋa [ɲi-gi qəsji ɣə-ɮde jə-ŋkʰə]   tə-ntɕʰu   \\
\\
\glt  I heard that he told you to come tomorrow.
\end{exe}

\begin{exe}
\ex
\gll ŋa [tʂaɕu ɲi-gi qəsji ɣə-ɮde jə-rə]   tə-ntɕʰu   \\
\\
\glt  I heard Bkrashis telling you to come tomorrow.
\end{exe}



\item smell \ipa{nənɔ}
\begin{exe}
\ex
\gll ŋi [pjenɔ tə-tsə-sə] de tə-nənɔ \\ %[pjenɔ tə-tsə-sə]-ʁa   tə-snəsnõ 
\\
\glt  I smelled that the meat was rotten.
\end{exe}

 
\begin{exe}
\ex
\gll   ŋi [tʂaɕi-ʁa ʁrəʁra]   tə-nənɔ \\ %[pjenɔ tə-tsə-sə]-ʁa   tə-snəsnõ 
\\
\glt  I smelled Bkrashis's smell.
\end{exe}

\begin{exe}
\ex
\gll    ŋi [tʂaɕu nə-ma-rʁirʁa-sə] de tə-nənɔ \\
\\
\glt  I smelled that Bkrashis did not have a bath.
\end{exe}

\item show \ipa{sni}
\begin{exe}
\ex
\gll    tu ŋa-gi pɔju ətɕəχi ra fɕi ŋə-rə kə-snã \\
\\
\glt  He showed me how to write in Tibetan.
\end{exe}
\begin{exe}
\ex
\gll    tu ŋa-gi ədəqʰej qiche kə-vrə-sə de kə-snã \\
\textsc{3sg:erg} \textsc{1sg-dat} \textsc{refl.gen}
\\
\glt  He showed me the new car the he bought.
\end{exe}

\item realize, know \ipa{kɔ}

\begin{exe}
\ex
\gll [ɲi jɔ tə-spej-sə] ŋa χavdə demdi χa ma-gõ \\
\\
\glt  I just noticed that you have moved.
\end{exe}

\begin{exe}
\ex
\gll [tə pəɕi ɣə-ɮdi-sə] ŋa   χa kõ \\
\\
\glt  I know that he has come this morning.
\end{exe}

\begin{exe}
\ex
\gll [rəsɲəske fɕe ke rka ŋɔ] ŋa   χa kõ \\
\\
\glt  I know that it is difficult to speak Rtau.
\end{exe}

\item think \ipa{ntsʰə}

\begin{exe}
\ex
\gll ŋa ɲi emtɕʰɔ tə-ɴqʰi-sə ntsʰu-rə \\
\\
\glt   I think that you have 减肥 a little
\end{exe}

\begin{exe}
\ex
\gll ŋa [tu jɔ tə-spɔ] ntsʰu-rə \\
\\
\glt   I think that he has moved.
\end{exe}
 
\begin{exe}
\ex
\gll tu [ŋa  tə-se] ntsʰə-rə \\
\\
\glt  He thinks that I died.
\end{exe}


\begin{exe}
\ex
\gll tu [ŋow jɔ  tə-spɔ] ntsʰə-rə \\
\\
\glt   I think that I have moved.
\end{exe}

\item believe, be  disappointed \ipa{jə tɕʰe}
 
\begin{exe}
\ex
\gll [tu  ŋa-gi  jə-lə] de ŋa jə tɕʰã-rə \\
\\
\glt  I believe what he told me. / Il m'a déçu.
\end{exe}

\begin{exe}
\ex
\gll [tu  liska nə-və-sə] de-gi   ŋa jə tɕʰã-rə \\
\\
\glt  I was disappointed by what he did.
\end{exe}

\item 受不了 \ipa{ʁmə}

\begin{exe}
\ex
\gll [tu  ŋa de zbəcʰa-cʰa]-kʰa   ŋa tə-ʁmõ \\
\\
\glt  I cannot stand being beaten by him anymore.
\end{exe}


 \item hope \ipa{riva tə}
 
 \begin{exe}
\ex
\gll [qəsji tə ɮde-lə] jə riva tə
\\
\glt  I hope that he will come tomorrow.
\end{exe}

 
 \begin{exe}
\ex
\gll [tə qəsji  ɮde-lə] ŋi riva tə
\\
\glt  I hope that he will come tomorrow.
\end{exe}

 \begin{exe}
\ex
\gll  ɲəqhej mətsʰi-nə riva zi cʰe de ətɕəgə ŋɔ ?\\
\\
\glt  I hope that he will come tomorrow.
\end{exe}

%\begin{exe}
%\ex
%\gll ŋa liska zɟəmɟu ntsʰə-sə cã\\
%\\
%\glt   I am thinking of changing my job.
%\end{exe}
\item dream \ipa{rɲə}


\begin{exe}
\ex
\gll ŋa [fɕupə tə-kʰã-sə] nə-rɲu \\
\\
\glt  I dreamt that I had become rich.
\end{exe}

\item remember \ipa{tʂẽ}
\begin{exe}
\ex
\gll ŋa tə xsɔ-fku  ɣɮdə-fku ŋə kə-dʑõ-sə de rə-dʐẽ \\
\\
\glt  I just remembered that we had met three or four years ago.
\end{exe}

\begin{exe}
\ex
\gll ŋa rəra ɕə-lə de rə-dʐẽ  \\
\\
\glt  I   remembered that I had to go shopping.
\end{exe}

\item hesitate \ipa{doˠpa  }
\begin{exe}
\ex
\gll ŋa tə-gi tʂõ-fɕe və-lə de toˠpa zã-rə \\
\\
\glt  I hesitate to tell him the truth.
\end{exe}

\item forget
\begin{exe}
\ex
\gll [ŋa ɕoˠdu zɔ-lə] tə-rmu-sə  \\
\\
\glt   I forgot my umbrella.
\end{exe}

\begin{exe}
\ex
\gll ŋaɲə pəsɲə tɕə fɕi-lə de ŋa tə-rmu-sə \\
  \\
\glt   I had forgotten that we had to meet today.
\end{exe}

\begin{exe}
\ex
\gll  [ŋa lɔvzõ-gi jɔ spɔ-ʁɔ kə-ʁow-sə] de  tə-rmə-sə.\\
  \\
\glt   Lobzang forgot that I had helped him to move his house.
\end{exe}
\item suspect \ipa{ftsu}

\begin{exe}
\ex
\gll ŋa [lɔvzõ-tɕʰa rgɔ rkə] kə-ftsu \\
  \\
\glt  I suspect that Lovzang stole the cow.
\end{exe} 

%\begin{exe}
%\ex
%\gll  tsə nə-nɔ-rə \\
%\\
%\glt  There is a rotting smell.
%\end{exe}

\end{enumerate}

\section{  verbs}



\end{enumerate}

\end{document}